\thetitle

### 2023-04-17 metatheory-condensed.tex
Removing mentioning of the appendix.
Move Why TyTkg into main text.

@@

%a \textbf{pen-and-paper} formalization
%assuming the facility of quotient inductive-inductive
%types~\cite{nordvall2010inductive,altkap2016} and extensional equality
%(equality reflection), available as a supplemental file in a
%\textbf{(pseudo-)Agda syntax}. Our formalization is directly following
%Agda syntax and type-checking (equality-checking) is done manually,
%leaving traces in the comments. We are currently working on mechanizing
%it and relying on Agda's rewriting mechanism to replace our limited
%usage of equality reflection. 
%
% We also typeset a more complete set of rules than the space allows and
% present it in
% \ifreport \cref{sec:complete-fmltt}.
% \else a supplemental appendix.
% \fi

>>>>

% Deal with intuition
\noindentparagraph{Why $\TyTkg{\lsig}$?}

%As shown by the equality rule \ruleref{tyeq/pk/add},
%for a given linkage signature $\lsig \coloneq \LSigAdd{\cdot}{\cdot}{\cdot}$,
%the type $\TyTkg{\lsig}$ is structurally similar to the linkage type $\TyLkg{\lsig}$
%except that $\TyTkg{\lsig}$ is a dependent tuple type---%
%it packages the fields of $\lsig$ into a dependent tuple type.
%
%We have computed several examples in
%\cref{fig:stlc-linkage-typing}, where the reader should also notice the high
%similarity between $\TyTkg{\lsig}$ and $\TyLkg{\lsig}$.

$\TyTkg{\lsig}$ is needed in rules $\ruleref{lsig/add}$ and $\ruleref{l/add}$:
these rules have a premise  $\goodTerm{\Gm,x : \TyTkg{\lsig}}{\seal}{A}$
responsible for possibly hiding W-type signatures (in $\TyTkg{\lsig}$) behind $\cU$ (in $A$);
see $\seal_6$ in \cref{fig:stlc-linkage-typing} for an example.
Here, both $\TyTkg{\lsig}$ and $A$ are dependent tuple types rather than linkage types.

%Thus the mystery of $\TyTkg{\cdot}$ really lies in its necessity. $\TyTkg{\cdot}$ is indispensible because \emph{we cannot do abstraction with $\TyLkg{\cdot}$ alone}, which in turn is due to the fact that an abstraction $A$ inside $\ruleref{lsig/add}$ and $\ruleref{l/add}$ is \emph{usually a dependent tuple type}.

An alternative to $\TyTkg{\lsig}$ would be to use linkage types $\TyLkg{\cdot}$ in this premise.
However, this alternative is infeasible.
We illustrate using a simple example.
Consider the linkage signature $\lsig$ below (it is written using the record
syntax for illustration but should not be confused with records, as a later field
is universally quantified over a $\mathit{self}$ variable).
The job is to hide the W-types $\TyS{\wcode{\wsig_1}}$ and
$\TyS{\wcode{\wsig_2}}$ in $\lsig$ behind $\cU$ in $\lsig^A$:
%
\begin{align*}
  \lsig &\coloneq \{ T_A : \TyS{\wcode{\wsig_1}}; &&T_B : \TyS{\wcode{\wsig_2}}; &&f : \El{\getfield{\mathit{self}}{T_A}} \to \El{\getfield{\mathit{self}}{T_B}} \} \\
  \lsig^A &\coloneq \{T_A : \cU; &&T_B : \cU; &&f : \El{\getfield{\mathit{self}}{T_A}} \to \El{\getfield{\mathit{self}}{T_B}} \} \\
  & && \goodTerm{x : \TyLkg{\lsig}}{\seal}{\TyLkg{\lsig^A}}
\end{align*}

%We illustrate with a small example using named convention, say given two inductive type $\wsig_1$ and $\wsig_2$ and a function $f$ maps from the first to the second type. Then we will have a linkage $\goodTerm{\cdot}{\ell}{\TyLkg{\lsig}}$. Say we want to add a new field into this linkage, and in $\ruleref{l/add}$, we chose $A$ to abstract both inductive type to $\cU$, resulting something like $\lsig^A$. The question now is whether $A$ is a linkage type $\TyLkg{\lsig^A}$ or a dependent tuple type $\TyTkg{\lsig^A}$; and for each case we will intuitively choose the following approach to do abstraction (choose $s$).

The problem lies in that the type $\TyLkg{\lsig^A}$ above cannot be inhabited.
Suppose there is some $\lkg^A$ such that $\cdot \vdash \lkg^A : \TyLkg{\lsig^A}$.
Then it follows from the self-parameterization that projecting out the last field~$f$
from $\lkg^A$ (notated as $\lkg^A.f$ for illustration) has the typing
$\goodTerm{T_A : \cU, T_B: \cU}{\lkg^A.f}{\El{T_A} \to \El{T_B}}$.
This typing means that there is a function of type
$\nTyPi{T_A}{\cU}{\nTyPi{T_B}{\cU}{\El{T_A}
\to \El{T_B}}}$, i.e., a function from an arbitrary type to another arbitrary
type. We can easily derive inconsistency from this function:
applying it to $\top$ and $\bot$ yields a function of type $\top\to\bot$.

%\begin{align*}
%  & \text{doing abstraction (choosing appropriate $s$ in \ruleref{l/add})} \\ 
%  \text{\textcolor{red}{Approach $\alpha$}}  \quad&\quad {\color{red} \TyLkg{\lsig} \longrightarrow \TyLkg{\lsig^A}} \\
%  \text{Approach } \beta \quad&\quad  \TyLkg{\lsig} \longrightarrow \TyTkg{\lsig} \longrightarrow \TyTkg{\lsig^A}\\
%\end{align*}
%
%However, the seemingly easier approach $\alpha$ is impossible because linkage
%$\TyLkg{\lsig^A}$ is uninhabited---%
%say we have $\ell^A : \TyLkg{\lsig^A}$, then by projection we will have an impossible term $\goodTerm{T_A : \cU, T_B: \cU}{\ell^A.f}{\El{T_A} \to \El{T_B}}$ due to the nature of overridability of linkage. To make the inconsistency clearer, by double abstraction we will have an impossible function $g \coloneq \goodTerm{\cdot}{\nlam{T_A}{\nlam{T_B}{\ell^A.f}}}{\nTyPi{T_A}{\cU}{\nTyPi{T_B}{\cU}{\El{T_A} \to \El{T_B}}}} $, i.e. a function maps from arbitrary type to another arbitrary type . We can easily derive inconsistency by applying $\napp{\napp{g}{\top}}{\bot} : \top \to \bot$. 
%
%Thus we can only choose the approach $\beta$ instead, which justifies why we have $\TyTkg{\cdot}$ in our typing rule, as $\TyTkg{\lsig}$ is usually the first step to do abstraction.

Thus, it is critical that the hiding of W-type signatures operate on dependent
tuple types rather than linkage types.
To this end, we introduce the syntax $\TyTkg{\lsig}$ and its computation rules
(e.g., \ruleref{tyeq/pk/add}) to make it convenient to package linkage signatures
into dependent tuple types.

%To summarize, the reason we need dependent tuple type is because of the overridability nature of

@@

%a \textbf{pen-and-paper} formalization
%assuming the facility of quotient inductive-inductive
%types~\cite{nordvall2010inductive,altkap2016} and extensional equality
%(equality reflection), available as a supplemental file in a
%\textbf{(pseudo-)Agda syntax}. Our formalization is directly following
%Agda syntax and type-checking (equality-checking) is done manually,
%leaving traces in the comments. We are currently working on mechanizing
%it and relying on Agda's rewriting mechanism to replace our limited
%usage of equality reflection. 
%
% We also typeset a more complete set of rules than the space allows and
% present it in
% \ifreport \cref{sec:complete-fmltt}.
% \else a 
% \fi
